A Large-scale Study of Polymorphism in R

### 2019-03-18 paper/main.tex
intro + r lang + bib

\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings}
\lstset{language=R}
\usepackage{my_style}

@@

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in over X millions
of lines of R code spread among some 7,500 packages, written over a period of
over 20 years by over 10,000 programmers.  We perform a dynamic analysis,
leveraging tests and use-cases, to determine the level of polymorphism that
is present in the code. We do this for several potential notions of
types. Our result suggest that polymorphism is important in some key parts
of the system but that relatively simple type annotations could be used to
capture most of the interesting cases.
\end{abstract}

@@

One way or another, R is here to stay, so we may as well figure out how to
deal with it.  One aspect of this is understanding R programmers and how
they interact with the language, and a step towards this goal is to
understand \AT{how types arise in R programs.}  We will shed some light on
this by analyzing the usage of R functions, and build a picture of how
polymorphic R code is.

\begin{itemize}
    \item Understanding R can help us improve the language, as well as
      better equip us to build languages which have an appeal outside of the
      realm of computer science.
    \item We're gonna tackle this by looking at how R programmers use types
      and type information in their programs.
    \item We will be mindful of the runtime representations of values to
      give more context to the information we glean from programmer-visible
      type information.
\end{itemize}

The main contributions of this work are:

\begin{itemize}
\item \AT{maybe} a tool to analyze the polymorphicity of R code;
\item a study of polymorphism in the R language \AT{using said tool};
\item a number of proposed static type systems for R which cover most of the
  type usage patterns which arise in R code.
\end{itemize}

%
\section{The R Programming Language}
\label{sec:rlang}

R is a lazy, side-effecting, dynamically-typed, ... .

Typical usage of the language is ... . \AT{Do we have something to cite for
  this?}  \AT{Should we motivate why we only look at R packages?  Some
  people might wonder why we don't look at non-library code.}

Most R libraries (called \textit{packages}) are stored in the Comprehensive
R Archive Network (CRAN) and Bioconductor package repositories.  \AT{Talk
  about important packages? tidyverse?}

\AT{Kind of ... want to combine the above two paragraphs somehow? Together,
  they lead in nicely to the genthat talk.}

\subsection{The \texttt{genthat} Package}

@@

Interestingly, we are somewhat able to replicate this using only package
code.  CRAN, has a policy that packages made available on the
CRAN platform should be accompanied by a series of examples, tests, and/or
vignettes showing off how the code is intended to be used.  Nominally
unrelated, the {\tt genthat} project~\cite{kvrikava2018tests} traces this swath of example
code to generate new examples which package designers may not have thought
of.  As it happens, we can leverage {\tt genthat} to extract dynamic type
information from running package test code, which presumably indicates
how package code is intended to be used.

@@

%
%
%
%
\subsection{Data Collection Pipeline}

@@

\begin{enumerate}

	\item first, all functions in {\tt p} are instrumented with on-exit hooks.
	These hooks call a number of R's available reflection functions on each of the arguments, as well as the return type:
	The called functions are {\tt typeof}, {\tt class}, and {\tt attributes}.
	See Section~\ref{sec:rlang} for details on these functions.

	\item then, depending on the specified type

\end{enumerate}

@@

%
%
%
%
\subsection{Attributes}
\label{sec:method:attributes}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio}

>>>>

\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings,xspace}
\lstset{language=R}
\usepackage{my_style}

@@

\newcommand{\PACKAGES}{7,500\xspace}
\newcommand{\PROGRAMMERS}{10,000\xspace}

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in over X millions
of lines of R code spread among some \PACKAGES packages, written over a
period of over 20 years by over 10,000 programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our result suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.
\end{abstract}

@@

We propose to capture the degree of polymorphism present in R by the means
of a dynamic analysis of widely used libraries. For each function call we
can record the types of its arguments and of its return value. This allows
us to observe how many different combination of types are accepted by any
given function. Unlike many other languages, R has a carefully curated
software repository called CRAN. To be deposited in CRAN, a package must
come with sample dataset, tests and executable use-cases. As part of normal
operations these tests are run regularly and failing packages are removed.
This allowed us to have access to \PACKAGES libraries and about an order of
magnitude more runnable scripts that exercise those libraries.

\section{The R Programming Language}\label{sec:rlang}

Over the last decade, the R Project has become a key tool for implementing
sophisticated data analysis algorithms in fields ranging from Computational
Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
the R project is a \emph{dynamic, lazy, functional, object-oriented}
programming language with a rather unusual combination of
features~\cite{ecoop12} designed to ease learning by non-programmer and
enable rapid development of new statistical methods.  The language, commonly
referred to as R was designed in 1993 by Ross Ihaka and Robert
Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
1995, under a GNU license, R rapidly became the lingua franca for
statistical data analysis. Today, there are over 13,000 R packages available
from repositories such as CRAN and Bioconductor.  With 55 R user groups
world-wide, Smith~\cite{eco11} estimates that there are over 2 million
end-users.

\section{The \texttt{genthat} Package}

@@

Interestingly, we are somewhat able to replicate this using only package
code.  CRAN, has a policy that packages made available on the CRAN platform
should be accompanied by a series of examples, tests, and/or vignettes
showing off how the code is intended to be used.  Nominally unrelated, the
{\tt genthat} project~\cite{issta18} traces this swath of example code to
generate new examples which package designers may not have thought of.  As
it happens, we can leverage {\tt genthat} to extract dynamic type
information from running package test code, which presumably indicates how
package code is intended to be used.

@@

%
\subsection{Data Collection Pipeline}

@@

\begin{enumerate}
\item first, all functions in {\tt p} are instrumented with on-exit hooks.
  These hooks call a number of R's available reflection functions on each of
  the arguments, as well as the return type: The called functions are {\tt
    typeof}, {\tt class}, and {\tt attributes}.  See Section~\ref{sec:rlang}
  for details on these functions.

\item then, depending on the specified type 
\end{enumerate}

@@

%
%
\subsection{Attributes}\label{sec:method:attributes}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
